Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-23 paper.Rnw
Fixed up all plots

Of particular relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing \multiatlas{} segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic \multiatlas{} strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using \multiatlas{} techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against \gs{} segmentations. %TODO: which segmentations were used? SNT?

@@

With at least five templates, \mb{} shows a mean improvement in similarity score
over \multiatlas{} segmentation when using the same size of atlas library and
majority vote label fusion (Figure \ref{fig:ADNI1-xval-k-diff}). The magnitude
of improvement increases with the size of template library, and shows
diminishing returns with larger atlas libraries. Peak improvement
(+\Sexpr{round(snt.mb.diff.best, digits=3)} DSC) is found with a single atlas
and template library of 19 images.

@@

%the CA1, CA4/DG, and SR/SL/SM
%subfields all show a percent error in volume that is at or lower than
%resampling error. The Subiculum and CA2/CA3 subfields show distinctly larger
%percent error in volume than is found through resampling.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: compare resampling
\begin{figure}[h!]
<<WAval-vol-boxplot,cache=T,fig.width=6,fig.height=4>>=
wa.xval.error = ddply(wa.xval.casted, c("file", "region"), function(df) {
  with(df, 
  data.frame(
    resampled = (gold_0.9mm - gold_0.3mm) / gold_0.3mm * 100, 
    mb_bravo  = (mb_bravo   - gold_0.3mm) / gold_0.3mm * 100, 
    mb_0.9mm  = (mb_0.9mm   - gold_0.3mm) / gold_0.3mm * 100))})

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: bland-altman 
\begin{figure}[h!]
<<WAval-bland,cache=T,fig.width=7,fig.height=6>>=
wa.xval.volumes = melt(wa.xval.casted, id.vars=c('file', 'region', 'gold_0.3mm'))
wa.xval.volumes = rename(wa.xval.volumes, c("variable"="measure", "value"="volume"))
levels(wa.xval.volumes$measure) <- list("Subsampled atlas labels"="gold_0.9mm",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

wa.xval.volumes$diff = wa.xval.volumes$gold_0.3mm - wa.xval.volumes$volume 
wa.xval.volumes$mean = (wa.xval.volumes$gold_0.3mm + wa.xval.volumes$volume)/2
wa.xval.bland.limits = ddply(wa.xval.volumes,c("region"),function(df) {
  data.frame(y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T))
})

@@

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point(size=1) + 
    geom_smooth(method="lm") + 
    xlab(expression("\\snt mean hippocampus volume " (mm^3))) + 
    ylab(expression("Automated mean hippocampal volume " (mm^3))) +
    annotate("text", y = c(2000,1500,1000,500), x=Inf-100, size=2.5,
              label = c(paste("FS r =", round(cor(df$SNT, df$FS),2)),
                        paste("FSL r =", round(cor(df$SNT, df$FSL),2)),
                        paste("MAPER r =", round(cor(df$SNT, df$MAPER),2)),
                        paste("MAGeT r =", round(cor(df$SNT, df$MAGeT),2))),
              hjust=1.1,vjust=1.5) +
    theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset{}
dataset by each of the four automated methods investigated (\freesurfer{} (FS), \fsl{},
\maper{}, \mb{}) vs. \snt{}. Linear fit lines and Pearson correlations are shown
for each method.}
  \label{fig:ADNI-volumes-plot}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Tabulation of segmentation accuracies of existing methods (ADNI)
%
%
\begin{table}[!tbp]
\scriptsize
\caption{\textbf{Automated segmentation accuracy (overlap with \snt{} labels)
of the ADNI dataset.} For each method, the number of labelled atlases
used for training, the best Dice's overlap measure, the disease classes
measured, and the validation procedure are shown. Unless specified, validation
datasets are composed equally of subjects diagnosed with Alzheimer's Disease
(AD), Mild Cognitive Impairment (MCI), and Cognitively Normal (CN). LOOCV =
Leave-one-out cross-validation. 
   Some studies of automated segmentation of ADNI images are excluded because they 
   do not not provide overlap measures for the hippocampus \citep{Heckemann2011,Chupin2009}.
  \label{tab:other-methods}} 
\begin{center}
\begin{tabular}{ l  c  p{1in}  l  p{2in}  }
\hline\hline
Method & Atlases & DSC & Reference & Validation \tabularnewline
\hline
LEAP                    &30&0.848                 &\cite{Wolz2010}  &Segmentation of 60 subjects\tabularnewline
ACM (AdaBoost-based)    &21&0.862                 &\cite{Morra2008} &LOOCV on atlases\tabularnewline
\mb{}                   &9 & \Sexpr{round(mb.kappa.best, digits=3)} & & 10
rounds of Monte Carlo CV on a pool of 69 subjects \tabularnewline % TODO: update
Patch-based label fusion&16&0.883 (CN) \newline 0.838 (AD)&\cite{Coupe2011e}&LOOCV on atlases\tabularnewline
Multi-atlas             &30&0.885                 &\cite{Lotjonen2010}&Segmentation of 60 subjects\tabularnewline
Multi-atlas + weighted fusion&20&0.898 (CN) \newline 0.798 (left HC, MCI) &\cite{Wang2011}  &10 rounds of Monte Carlo CV on 20 subjects, pool of 139 (CN/MCI)\tabularnewline
Multi-atlas (MAPS)    &55&0.890                   &\cite{Leung2010}&Segmentation of 30 subjects (10 AD, MCI, and CN)\tabularnewline
\hline
\end{tabular}
\end{center}
\end{table}

>>>>

Of particular relevance to the present work is the LEAP algorithm (Learning
Embeddings for Atlas Propagation; \citet{Wolz2010}) because of its focus on
performing \multiatlas{} segmentation with a limited number of input atlases.  The
LEAP algorithm is a clever modification to the basic \multiatlas{} strategy in
which an atlas library is grown, beginning with a set of manually labelled
atlases, by successively incorporating unlabelled target images once they
themselves have been labelled using \multiatlas{} techniques. The sequence in
which target images are labelled is chosen so that the similarity between the
atlas images and the target images is minimised at each step, effectively
allowing for deformations between very dissimilar images to be broken up into
sequences of smaller deformations. Although \citet{Wolz2010} begin with an atlas
library of 30 MR images, this method could theoretically work using a much
smaller atlas library.  In their validation, LEAP was used to segment the whole
hippocampus in the ADNI-1 baseline dataset, achieving a mean Dice score of 0.85
against semi-automated segmentations.

@@

With at least five templates, \mb{} shows a mean improvement in similarity score
over \multiatlas{} segmentation when using the same size of atlas library and
majority vote label fusion (Figure \ref{fig:ADNI1-xval-k-diff}). The magnitude
of improvement increases with the size of template library, and shows
diminishing returns with larger atlas libraries. Peak improvement
(+\Sexpr{round(mb.diff.best, digits=3)} DSC) is found with a single atlas
and template library of 19 images.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: compare resampling
\begin{figure}[h!]
<<WAval-vol-boxplot,cache=T,fig.width=6,fig.height=5>>=
wa.xval.error = ddply(wa.xval.casted, c("file", "region"), function(df) {
  with(df, 
  data.frame(
    resampled = (gold_0.9mm - gold_0.3mm) / gold_0.3mm * 100, 
    mb_bravo  = (mb_bravo   - gold_0.3mm) / gold_0.3mm * 100, 
    mb_0.9mm  = (mb_0.9mm   - gold_0.3mm) / gold_0.3mm * 100))})

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: bland-altman 
\begin{figure}[h!]
<<WAval-bland,cache=T,fig.width=7,fig.height=6>>=
wa.xval.volumes = melt(wa.xval.casted, id.vars=c('file', 'region', 'gold_0.3mm', 'gold_0.9mm'))
wa.xval.volumes = rename(wa.xval.volumes, c("variable"="measure", "value"="volume"))
levels(wa.xval.volumes$measure) <- list("Subsampled atlas labels"="gold_0.9mm",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

wa.xval.volumes$diff = wa.xval.volumes$gold_0.9mm - wa.xval.volumes$volume 
wa.xval.volumes$mean = (wa.xval.volumes$gold_0.9mm + wa.xval.volumes$volume)/2
wa.xval.bland.limits = ddply(wa.xval.volumes,c("region"),function(df) {
  data.frame(y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T))
})

@@

adni.volumes.lm = ddply(melted,c("Method"), function (df) {
  data.frame(eqn=lm_eqn(lm(volume ~ SNT, df)))})
adni.volumes.lm$y = 6450-0:3*300

ggplot(data=melted, aes(x=SNT, y=volume, colour=Method)) + 
    geom_point(size=1) + 
    geom_smooth(method="lm") + 
    xlab(expression("SNT mean hippocampus volume " (mm^3))) + 
    ylab(expression("Automated mean hippocampal volume " (mm^3))) +
    geom_text(aes(x=850,y=y,label=eqn),parse=T,data=adni.volumes.lm,
      size=2.5,hjust=0) +
    theme(legend.direction = "horizontal", legend.position = "bottom")

#    annotate("text", y = c(2000,1500,1000,500), x=Inf-100, size=2.5,
#              label = c(paste("FS r =", round(cor(df$SNT, df$FS),2)),
#                        paste("FSL r =", round(cor(df$SNT, df$FSL),2)),
#                        paste("MAPER r =", round(cor(df$SNT, df$MAPER),2)),
#                        paste("MAGeT r =", round(cor(df$SNT, df$MAGeT),2))),
#              hjust=1.1,vjust=1.5) +
@
  \caption{Subject mean hippocampal volume as measured in the \adnidataset{}
dataset by each of the four automated methods investigated (\freesurfer{} (FS), \fsl{},
\maper{}, \mb{}) vs. \snt{}. Linear fit lines and Pearson correlations are shown
for each method.}
  \label{fig:ADNI-volumes-plot}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Tabulation of segmentation accuracies of existing methods (ADNI)
%
%
\begin{table}[!tbp]
\scriptsize
\caption{\textbf{Automated segmentation accuracy (overlap with \snt{} labels)
of the ADNI dataset.} For each method, the number of labelled atlases
used for training, the best Dice's overlap measure, the disease classes
measured, and the validation procedure are shown. Unless specified, validation
datasets are composed equally of subjects diagnosed with Alzheimer's Disease
(AD), Mild Cognitive Impairment (MCI), and Cognitively Normal (CN). LOOCV =
Leave-one-out cross-validation. 
   Some studies of automated segmentation of ADNI images are excluded because they 
   do not not provide overlap measures for the hippocampus \citep{Heckemann2011,Chupin2009}.
  \label{tab:other-methods}} 
\begin{center}
\begin{tabular}{ l  c  p{1in}  l  p{2in}  }
\hline\hline
Method & Atlases & DSC & Reference & Validation \tabularnewline
\hline
LEAP                    &30&0.848                 &\cite{Wolz2010}
&Segmentation of 182 subjects\tabularnewline
ACM (AdaBoost-based)    &21&0.862                 &\cite{Morra2008} &LOOCV on atlases\tabularnewline
\mb{}                   &9 & \Sexpr{round(mb.kappa.best, digits=3)} & & 10
rounds of Monte Carlo CV on a pool of 69 subjects \tabularnewline % TODO: update
Patch-based label fusion&16&0.883 (CN) \newline 0.838 (AD)&\cite{Coupe2011e}&LOOCV on atlases\tabularnewline
Multi-atlas             &30&0.885                 &\cite{Lotjonen2010}&Segmentation of 60 subjects\tabularnewline
Multi-atlas + weighted fusion&20&0.898 (CN) \newline 0.798 (left HC, MCI) &\cite{Wang2011}  &10 rounds of Monte Carlo CV on 20 subjects, pool of 139 (CN/MCI)\tabularnewline
Multi-atlas (MAPS)    &55&0.890                   &\cite{Leung2010}&Segmentation of 30 subjects (10 AD, MCI, and CN)\tabularnewline
\hline
\end{tabular}
\end{center}
\end{table}
